Modelling with ANIMO:\\ between Fuzzy Logic and Differential Equations

### 2015-05-20 Paper.tex
Updated models and figures for TNF-EGF and Circadian clock

\pgfdeclareimage[height=3ex]{win-logo}{images/win_logo}
\pgfdeclareimage[height=3ex]{mac-logo}{images/mac_logo}
\pgfdeclareimage[height=3ex]{linux-logo}{images/linux_logo}
\pgfdeclareimage[height=3ex]{ubuntu-logo}{images/ubuntu_logo}
\pgfdeclareimage[height=2ex]{mac-command}{images/mac_command}
\pgfdeclareimage[height=2ex]{lower-bound}{images/lowerbound}
\pgfdeclareimage[height=2ex]{upper-bound}{images/upperbound_green}

@@

\title{Modelling with ANIMO:\\ between Fuzzy Logic and ODE}%ODE vs Fuzzy Logic: in medio stat ANIMO}%Formal models of biological networks in \animo2{}} 
%Building formal models of biological networks avoiding formal training}

@@

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
The large amount of data an experimental biologist
must deal with makes it nearly impossible to reason on
the dynamics of complex systems without relying on computational
support.
The software tool ANIMO (Analysis of Networks with Interactive MOdelling)
provides such computational support and allows insight into the complex networks of signaling events occurring in living cells.
The user-friendly interface allows biology experts to exploit formal methods
without the need to become experts also in mathematical formalisms.
\parttitle{Results}
ANIMO models show a degree of precision intermediate between ordinary differential equations (ODEs) and fuzzy logic.
We show this by presenting ANIMO models for two case studies: \emph{Drosophila melanogaster} circadian clock, and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells.
The models were originally developed with ODEs and
fuzzy logic respectively.
\parttitle{Conclusions}
% The software tool ANIMO, recently updated to work with the latest version of Cytoscape,
% allows the biologist to develop a formal model through a user friendly interface,
% achieving useful insight without the need to study any mathematical formalism.
%
ANIMO models replicate with good precision the results of both the ODE and
fuzzy logic models. Moreover, ANIMO models require less parameters than ODEs and are more precise than
fuzzy logic. For this reason we position the modelling paradigm of ANIMO
between ODEs and fuzzy logic.
\end{abstract}

@@

\section*{Background}\label{sec:introduction}
\subsection*{Modelling in cell biology}
Executable biology is a young subfield in computational modeling, aimed at constructing models that mimic
biological phenomena \emph{in silico}. It provides an interesting paradigm for enhancing network
diagrams with an underlying formal description of both network components and their interactions.
Executable models typically contain descriptions of the direct interactions of network components,
requiring the modeler to think in terms of
``cause and effect''. Biologists reason in similar ways about the molecular mechanisms of network
interactions, and this makes construction of an executable model an intuitive process.
ANIMO (Analysis of Networks with Interactive MOdeling,~\cite{animo-ieee,animo-gene})
provides an enabling technology to increase the use of computational models by experimental
 biologists without a thorough mathematical background. Such models are indispensable for formally
comparing experimental data to prior knowledge, or for structuring experimental findings into a
new theory.
When dealing with complex biological networks executable biology models are particularly useful
to understand the non-linear dynamics and the entailed emergent properties of the networks.
In those cases, an ANIMO model can be used as a support to obtain insight based on abstract
representations of the interactions occurring inside living cells.
Other applications of ANIMO models include performing \emph{in silico} experiments and
the storage and transfer of knowledge on biological networks.

\subsection*{A brief introduction to ANIMO}
The user interface of ANIMO is displayed in Figure~\ref{fig:cytoscape},
where we present an example of biological network enriched with dynamic information.
ANIMO is implemented as a plug-in to Cytoscape~\cite{cytoscape}, a software tool developed to represent biological networks.
On top of the static topological information displayed in Cytoscape, ANIMO represents
biological interactions starting from the basic concept of \emph{activity}:
each biological entity in an ANIMO model is considered to be either active or inactive.
Activity is to be intended in a very broad sense: for example, an active gene is being transcribed,
an active kinase can perform phosphorylations, etc. Each node in an ANIMO network represents
both active and inactive entities of the same type, with the relative amount of active entities
represented by the node color on a user-configurable scale. Interactions among nodes define
how the biological entities in a network influence each other's activity.
Only nodes whose activity is larger than 0 can have an 
influence on their downstream targets, and only if that influence is not counterbalanced by intervening opposite interactions.
For example, the interaction $A\rightarrow B$ (read ``A activates B'') indicates that node A, if active,
will increase the activity of node B. If we add an additional interaction to the example,
$C \dashv$\hspace{0.1em}$B$ (``C inhibits B''), with C also active, then the activity of B will change depending on the
activity levels of A and C, and on the \emph{strength} of their influence.
The strength of an ANIMO interaction is a scale factor $k$ which defines the speed at which
that interaction occurs: higher values of $k$ give faster interactions. These $k$-values are
the only parameters needed in an ANIMO model, and can be given in either
a quantitative (as real numbers) or a qualitative way, choosing among self-explanatory descriptions
such as ``very slow'', ``slow'', ``medium'', ``fast'', ``very fast''.

\begin{figure*}[htbp]
\begin{center}
   \includegraphics[width=.85\textwidth]{images/screenshotTNFmodelSmall}
\end{center}
\caption{\csentence{The Cytoscape user interface running the ANIMO plug-in.}
The \emph{Network} panel in the center contains the nodes-edges
model of the example TNF$\alpha$ patwhay (see Methods Section), with
colors indicating node activity levels and shapes representing different protein categories (see the \emph{Legend} on the left).
The \emph{Results Panel} on the right contains a graph plotting activity levels of selected nodes
during the first 24 hours of evolution of the model. The slider under the graph
allows the user to select the time instant (marked as a vertical red line in the graph) on which
the colors nodes in the \emph{Network} are based.
The series with the {\sf \_data} suffix is experimental
data from~\cite{pathway-compendium}, considering a treatment with 100~ng/ml TNF$\alpha$.\\
All acronyms used in this paper
and their corresponding UniProt IDs are listed in Suppl. Sect.~\ref{suppl-sec:names}.
\label{fig:cytoscape}}
\end{figure*}

ANIMO models are analyzed to produce graphs showing how the activity of selected nodes
change with time, allowing the user to obtain a view on the dynamic behavior of their network.
In order to obtain these results, a model defined in ANIMO is automatically translated into
its corresponding representation as a network of Timed Automata~\cite{timed-automata-alur-dill}
and then analyzed \emph{behind the scenes} with the software tool UPPAAL~\cite{uppaal}.
The formal language of Timed Automata allows to represent and analyze complex behaviors precisely and efficiently,
but the user does not need to understand how Timed Automata or UPPAAL work, as the analysis process is made transparent.
A fastidious user can still access the underlying models and perform other analyses in UPPAAL,
but that is not required in order to fully profit from ANIMO.

@@

Figure~\ref{fig:animo-spectrum} shows the position of ANIMO in a spectrum of modeling methods: justification for this placement is given in the Results section.

\begin{figure*}[htbp]
 \begin{center}
  \includegraphics[width=0.9\textwidth]{images/modeling_methods_spectrum_animo}
 \end{center}
\caption{\csentence{The spectrum of modeling methods, with the addition of ANIMO.}
The precision of ANIMO models is halfway between fuzzy logic and ODE. In addition, ANIMO
allows for an easier modeling experience thanks to a user-friendly interface based on
the widely used network modeling software Cytoscape.\label{fig:animo-spectrum}}
\end{figure*}

@@

\section*{Results}\label{sec:results}
\subsection*{Modelling oscillation}\label{sec:animo-drosophila}
To demonstrate that results obtained with ANIMO are comparable to results from widely used modeling
approaches, we present 
an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster} (Fig.~\ref{fig:drosophila-model}).
This ANIMO model is based on~\cite{drosophila-ode-model},
where ordinary differential equations (ODEs) were used.
The cyclic behavior of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO acts as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~\ref{suppl:repressilator}).

@@

We constructed a model of the signaling network downstream of TNF$\alpha$ and EGF
in HT-29 human colo carcinoma cells, formalizing
the crosstalk that takes place between the pathways at different levels of cellular regulation.
We first modeled the two pathways in isolation~(Suppl. Figs.~\ref{fig:large-model-tnf}, \ref{fig:large-model-egf}),
using information on protein interactions from
the KEGG~\cite{kegg} and phosphosite~\cite{phosphosite} databases. These models were fitted to experimental data
from previous studies~\cite{pathway-compendium,pathway-autocrine}.
We then merged the two pathways into a single model and added autocrine crosstalk between the pathways that has been
described in~\cite{pathway-autocrine}.
Briefly, stimulation with TNF$\alpha$ ({\sf TNFa} in the model) leads to a rapid release of TGF$\alpha$ ({\sf TGFa}),
which activates the EGF receptor ({\sf EGFR}).
This activation causes secretion of IL-1$\alpha$ ({\sf IL-1a}) at later time points.
The effect of IL-1$\alpha$ is down-regulated by the secretion of IL-1 receptor antagonist ({\sf IL-1ra})
downstream of TNF$\alpha$.
The resulting model (Fig.~\ref{fig:large-model-all}A) was compared to the experimental data
for treatments with 100 ng/ml TNF alone and 100 ng/ml EGF alone (data not shown)~\cite{pathway-compendium}.

\def\largeModelScale{0.18}%0.15}%0.155}%0.27}
\def\legendScalaColori{0.21}%0.18}%0.21}
\def\legendScalaForme{0.21}%0.18}%0.21}
\def\scalaGrafici{0.0709}%0.13}
\begin{figure*}[!htpb]
\centering
%  \subfloat[\label{fig:large-model-no-hypotheses}]{\includegraphics[scale=\largeModelScale]{images/large_network_merged_no_hypotheses}}\\
% %\subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme}}\\ \subfloat{\includegraphics[scale=\legendScalaForme]{images/legenda_colori}}\\
% \subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme_e_colori}}\\
% \addtocounter{subfigure}{-1}
% \subfloat[\label{fig:large-model-graph1}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_TNF0_ho_hyp_IL-1a3}}
% \subfloat[\label{fig:large-model-graph3}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_no_hyp_JNK1_MK2}}
  \includegraphics[width=0.9\textwidth]{images/large_network_merged_no_hypotheses_oneImage}
\caption{\csentence{Signaling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-no-hypotheses}}
The model for the merged TNF$\alpha$ and EGF pathways. Node colors represent the
activity level of the corresponding modeled reactants at time $t = 15$ minutes after
a stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph1}}~
Modeled production of IL-1$\alpha$ after stimulation with 100 ng/ml TGF$\alpha$ (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph3}}~
Modeled activation of JNK1 and MK2 after stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-all}
\end{figure*}

At this point, the behavior of the model deviated from the data for some of the nodes.
Changing the parameters of the model was not enough to reproduce the behaviour shown by experimental
data. This is an interesting situation, as it requires changes in the topology of
the model~\cite{animo-syncop}, which can be interpreted as new hypotheses. Below, we give two examples and show how
adaptation of the model can be used to generate novel testable hypotheses.

@@

\begin{figure*}[!htpb]
\begin{center}
% \subfloat[\label{fig:large-model-complete}]{\includegraphics[scale=\largeModelScale]{images/large_network_legendg}}\\
%  \subfloat[\label{fig:large-model-graph2}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_vs_TNF100_hyp1_IL-1a4}}
%  \subfloat[\label{fig:large-model-graph4}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_hyp2_JNK1_MK2}}
  \includegraphics[width=0.9\textwidth]{images/large_network_legendg_oneImage}
\end{center}
\caption{\csentence{Signaling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells: improved model.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-complete}}
The model for the merged TNF$\alpha$ and EGF pathways
after addition of the two hypotheses (highlighted).
Hypothesis 1 assumes IL-1$\alpha$ expression to depend on AP-1 activity, which in turn requires
both c-Jun en c-Fos to be activated by JNK1 and ERK, respectively. Hypothesis 2 assumes RAS as an activator
of MEKK1. Node colors represent the activity levels $15$ minutes
after stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph2}}~
After the addition of the first hypothesis (activation of IL-1$\alpha$ production depending both
on JNK1 and ERK): production of IL-1$\alpha$ after stimulation with 100 ng/ml TNF$\alpha$ (series {\sf IL-1a~(TNFa)})
compared with stimulation with 100 ng/ml TGF$\alpha$ (series {\sf IL-1a~(TGFa)}) (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph4}}~
After the addition of the second hypothesis (activation of MEKK1 downstream of EGFR):
stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
Suppl. Sect.~\ref{suppl:parameters-tnf-egf} explains how the dosage of 5 ng/ml TNF$\alpha$ was represented in the model.\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-complete}
\end{figure*}

@@

\section*{Discussion}
\subsection*{Final remarks on the models}
We described the construction of an ANIMO
model of the circadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modeling abstractions on which ANIMO is based. The biggest
difference between the construction of these models is that the model in~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO a number of network
nodes is drawn for the molecules involved.
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter
that defines the strength of each
interaction. This is a more intuitive approach to construct a model.
Further contributing to an interactive modeling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.
All this comes at the price of lower model precision: the curves representing
oscillation of protein activities in the ANIMO model are not as smooth
nor as precise as those obtained from the original ODE model (see Suppl. Fig.~\ref{suppl-fig:grafici-drosophila}).

We also showed the construction of an executable model
of signaling events downstream of
TNF$\alpha$ and EGF in human colon carcinoma cells. This data set has been used for
previous modeling studies, based on partial least-squares regression and fuzzy
logic~\cite{pathway-leastsquare,pathway-fuzzy}.
The partial least-squares regression model describes an abstract data-driven model
that uses statistical correlations
to relate signal transduction events to various cellular decisions. This type of modeling is
very useful in uncovering new and unexpected relations. It is also successful in making
predictions, but gives little direct insight in the dynamic behavior of the network (see Fig.~\ref{fig:animo-spectrum}). Fuzzy
logic analysis led to a model that gave a better fit to the dynamic network behavior than
discrete logic (Boolean) models. Inspection of the inputs to the logical gates that were used
to model protein behavior led to the prediction of novel interactions between proteins,
showing the usefulness of this approach. For most of the proteins, such as JNK1, time was
used as an input parameter. For example, the logical gates ``if TNF$\alpha$ is high
\emph{AND} time is low, then JNK1 is high'' and ``if TNF$\alpha$ is high \emph{AND} time is
high, then JNK1 is low'' were used to
describe the dynamic behavior of JNK1. Although this leads to a representative
description of the dynamic behavior of JNK1, peaks in protein activity at early time points,
as measured in wet-lab experiments, were
not reproduced by the fuzzy logic model. Moreover, the fuzzy logic model gave no insight in the molecular interactions
that are involved in activation or inhibition.

>>>>

\pgfdeclareimage[height=3ex]{win-logo}{images/win_logo}
\pgfdeclareimage[height=3ex]{mac-logo}{images/mac_logo}
\pgfdeclareimage[height=3ex]{linux-logo}{images/linux_logo}
\pgfdeclareimage[height=3ex]{ubuntu-logo}{images/ubuntu_logo}
\pgfdeclareimage[height=2ex]{mac-command}{images/mac_command}
\pgfdeclareimage[height=2ex]{lower-bound}{images/lowerbound}
\pgfdeclareimage[height=2ex]{upper-bound}{images/upperbound_green}
\pgfdeclareimage[width=\textwidth]{100TNF0EGF}{images/TNFa-EGF/100TNFa-0EGF/100TNF_0EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=\textwidth]{0TNF100EGF}{images/TNFa-EGF/0TNFa-100EGF/0TNF_100EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=\textwidth]{100TNF100EGF}{images/TNFa-EGF/100TNFa-100EGF/100TNF_100EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=\textwidth]{CircadianComparison}{images/Circadian/ANIMOvsODE-crop}

@@

\title{Modelling with ANIMO:\\ between Fuzzy Logic and Differential Equations}%ODE vs Fuzzy Logic: in medio stat ANIMO}%Formal models of biological networks in \animo2{}} 
%Building formal models of biological networks avoiding formal training}

@@

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
The large amount of data an experimental biologist
must deal with makes the use of computational methods essential
to reason on the dynamics of complex systems.
The software tool ANIMO (Analysis of Networks with Interactive MOdelling)
provides such computational support and allows insight into the complex networks of signalling events occurring in living cells.
Biology experts are able to exploit the power of the underlying formal methods
via a user interface specifically tailored for biological applications.
\parttitle{Results}
ANIMO models show a degree of precision intermediate between ordinary differential equations (ODEs) and fuzzy logic.
We show this by presenting ANIMO models for two case studies: \emph{Drosophila melanogaster} circadian clock, and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells.
The models were originally developed with ODEs and
fuzzy logic respectively.
\parttitle{Conclusions}
% The software tool ANIMO, recently updated to work with the latest version of Cytoscape,
% allows the biologist to develop a formal model through a user friendly interface,
% achieving useful insight without the need to study any mathematical formalism.
%
ANIMO models replicate with good precision the results of both the ODE and
fuzzy logic models. Moreover, ANIMO models require less parameters than ODEs and are more precise than
fuzzy logic. For this reason we position the modelling paradigm of ANIMO
between ODEs and fuzzy logic.
\end{abstract}

@@

\section*{Background}\label{sec:introduction}
\subsection*{Modelling in cell biology}
Executable biology is a young subfield in computational modelling, aimed at constructing models that mimic
biological phenomena \emph{in silico}. It provides an interesting paradigm for enhancing network
diagrams with an underlying formal description of both network components and their interactions.
Executable models typically contain descriptions of the direct interactions of network components,
requiring the modeller to think in terms of
``cause and effect''. Biologists reason in similar ways about the molecular mechanisms of network
interactions, and this makes the construction of an executable model an intuitive process.
ANIMO (Analysis of Networks with Interactive MOdelling,~\cite{animo-ieee,animo-gene})
provides an enabling technology to increase the use of computational models by experimental
biologists using their domain-specific language. Such models are indispensable for formally
comparing experimental data to prior knowledge, or for structuring experimental findings into a
new theory.
When dealing with complex biological networks executable biology models are particularly useful
to understand the non-linear dynamics and the entailed emergent properties of the networks.
In those cases, an ANIMO model can be used as a support to obtain insight based on abstract
representations of the interactions occurring inside living cells.
Other applications of ANIMO models include performing \emph{in silico} experiments and
the storage and transfer of knowledge on biological networks.

\subsection*{A brief introduction to ANIMO}
The user interface of ANIMO is displayed in Figure~\ref{fig:cytoscape},
where we present an example of a biological network enriched with dynamic information.
ANIMO is implemented as a plug-in to Cytoscape~\cite{cytoscape}, a software tool developed to represent biological networks.
On top of the static topological information displayed in Cytoscape, ANIMO represents
biological interactions starting from the basic concept of \emph{activity}:
each biological entity in an ANIMO model is considered to be either active or inactive.
Activity is to be intended in a very broad sense: for example, an active gene is being transcribed,
an active kinase can perform phosphorylations, etc. Each node in an ANIMO network represents
both active and inactive entities of the same type, with the relative amount of active entities
represented by the node colour on a user-configurable scale. Interactions among nodes define
how the biological entities in a network influence each other's activity.
Only nodes whose activity is larger than 0 can have an 
influence on their downstream targets, and only if that influence is not counterbalanced by intervening opposite interactions.
For example, the interaction $A\rightarrow B$ (read ``A activates B'') indicates that node A, if active,
will increase the activity of node B. If we add an additional interaction to the example,
$C \dashv$\hspace{0.1em}$B$ (``C inhibits B''), with C also active, then the activity of B will change depending on the
activity levels of A and C, and on the \emph{strength} of their influence.
The strength of an ANIMO interaction is a scale factor $k$ which defines the speed at which
that interaction occurs: higher values of $k$ give faster interactions. These $k$-values are
the only parameters needed in an ANIMO model, and can be given in either
a quantitative (as real numbers) or a qualitative way, choosing among self-explanatory descriptions
such as ``very slow'', ``slow'', ``medium'', ``fast'', ``very fast''.

\begin{figure*}[htbp]
\begin{center}
   \includegraphics[width=.85\textwidth]{images/screenshotTNFmodelSmall}
\end{center}
\caption{\csentence{The Cytoscape user interface running the ANIMO plug-in.}
The \emph{Network} panel in the centre contains the nodes-edges
model of the example TNF$\alpha$ pathway (see Methods Section), with
colours indicating node activity levels and shapes representing different protein categories (see the \emph{Legend} on the left).
The \emph{Results Panel} on the right contains a graph plotting activity levels of selected nodes
during the first 24 hours of evolution of the model. The slider under the graph
allows the user to select the time instant (marked as a vertical red line in the graph) on which
the colours nodes in the \emph{Network} are based.
The series with the {\sf \_data} suffix is experimental
data from~\cite{pathway-compendium}, considering a treatment with 100~ng/ml TNF$\alpha$.\\
All acronyms used in this paper
and their corresponding UniProt IDs are listed in Suppl. Sect.~\ref{suppl-sec:names}.
\label{fig:cytoscape}}
\end{figure*}

ANIMO models are analysed to produce graphs showing how the activity of selected nodes
change with time, allowing the user to obtain a view on the dynamic behaviour of their network.
In order to obtain these results, a model defined in ANIMO is automatically translated into
its corresponding representation as a network of Timed Automata~\cite{timed-automata-alur-dill}
and then analysed \emph{behind the scenes} with the software tool UPPAAL~\cite{uppaal}.
The formal language of Timed Automata allows to represent and analyse complex behaviours precisely and efficiently,
but the user does not need to directly manipulate Timed Automata or UPPAAL, as the analysis process is made transparent.
A fastidious user can still access the underlying models and perform other analyses in UPPAAL,
but that is not required in order to fully profit from ANIMO.

@@

Figure~\ref{fig:animo-spectrum} shows the position of ANIMO in a spectrum of modelling methods: justification for this placement is given in the Results section.

\begin{figure*}[htbp]
 \begin{center}
  \includegraphics[width=0.9\textwidth]{images/modeling_methods_spectrum_animo}
 \end{center}
\caption{\csentence{The spectrum of modelling methods, with the addition of ANIMO.}
The precision of ANIMO models is halfway between fuzzy logic and ODE. In addition, ANIMO
allows for an easier modelling experience thanks to a user-friendly interface based on
the widely used network modelling software Cytoscape.\label{fig:animo-spectrum}}
\end{figure*}

@@

\section*{Results}\label{sec:results}
\subsection*{Modelling oscillation}\label{sec:animo-drosophila}
To demonstrate that results obtained with ANIMO are comparable to results from widely used modelling
approaches, we present 
an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster} (Fig.~\ref{fig:drosophila-model}).
This ANIMO model is based on~\cite{drosophila-ode-model},
where ordinary differential equations (ODEs) were used.
The cyclic behaviour of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO acts as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~\ref{suppl:repressilator}).

@@

We constructed a model of the signalling network downstream of TNF$\alpha$ and EGF
in HT-29 human colon carcinoma cells, formalizing
the crosstalk that takes place between the pathways at different levels of cellular regulation.
We first modelled the two pathways in isolation~(Suppl. Figs.~\ref{fig:large-model-tnf}, \ref{fig:large-model-egf}),
using information on protein interactions from
the KEGG~\cite{kegg} and phosphosite~\cite{phosphosite} databases. These models were fitted to experimental data
from previous studies~\cite{pathway-compendium,pathway-autocrine}.
We then merged the two pathways into a single model and added autocrine crosstalk between the pathways that has been
described in~\cite{pathway-autocrine}.
Briefly, stimulation with TNF$\alpha$ ({\sf TNFa} in the model) leads to a rapid release of TGF$\alpha$ ({\sf TGFa}),
which activates the EGF receptor ({\sf EGFR}).
This activation causes secretion of IL-1$\alpha$ ({\sf IL-1a}) at later time points.
The effect of IL-1$\alpha$ is down-regulated by the secretion of IL-1 receptor antagonist ({\sf IL-1ra})
downstream of TNF$\alpha$.
The resulting model (Fig.~\ref{fig:large-model-all}A) was compared to the experimental data
for treatments with 100 ng/ml TNF alone and 100 ng/ml EGF alone (data not shown)~\cite{pathway-compendium}.

\def\largeModelScale{0.18}%0.15}%0.155}%0.27}
\def\legendScalaColori{0.21}%0.18}%0.21}
\def\legendScalaForme{0.21}%0.18}%0.21}
\def\scalaGrafici{0.0709}%0.13}
\begin{figure*}[!htpb]
\centering
%  \subfloat[\label{fig:large-model-no-hypotheses}]{\includegraphics[scale=\largeModelScale]{images/large_network_merged_no_hypotheses}}\\
% %\subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme}}\\ \subfloat{\includegraphics[scale=\legendScalaForme]{images/legenda_colori}}\\
% \subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme_e_colori}}\\
% \addtocounter{subfigure}{-1}
% \subfloat[\label{fig:large-model-graph1}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_TNF0_ho_hyp_IL-1a3}}
% \subfloat[\label{fig:large-model-graph3}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_no_hyp_JNK1_MK2}}
  \includegraphics[width=0.9\textwidth]{images/large_network_merged_no_hypotheses_oneImage}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-no-hypotheses}}
The model for the merged TNF$\alpha$ and EGF pathways. Node colours represent the
activity level of the corresponding modelled reactants at time $t = 15$ minutes after
a stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph1}}~
Modelled production of IL-1$\alpha$ after stimulation with 100 ng/ml TGF$\alpha$ (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph3}}~
Modelled activation of JNK1 and MK2 after stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-all}
\end{figure*}

At this point, the behaviour of the model deviated from the data for some of the nodes.
Changing the parameters of the model was not enough to reproduce the behaviour shown by experimental
data. This is an interesting situation, as it requires changes in the topology of
the model~\cite{animo-syncop}, reflecting the formulation of a new hypothesis
on the structure of the model. Below, we give two examples and show how
adaptation of the model can be used to generate novel testable hypotheses.

@@

\begin{figure*}[!htpb]
\begin{center}
% \subfloat[\label{fig:large-model-complete}]{\includegraphics[scale=\largeModelScale]{images/large_network_legendg}}\\
%  \subfloat[\label{fig:large-model-graph2}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_vs_TNF100_hyp1_IL-1a4}}
%  \subfloat[\label{fig:large-model-graph4}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_hyp2_JNK1_MK2}}
  \includegraphics[width=0.9\textwidth]{images/large_network_legendg_oneImage}
\end{center}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells: improved model.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-complete}}
The model for the merged TNF$\alpha$ and EGF pathways
after addition of the two hypotheses (highlighted).
Hypothesis 1 assumes IL-1$\alpha$ expression to depend on AP-1 activity, which in turn requires
both c-Jun en c-Fos to be activated by JNK1 and ERK, respectively. Hypothesis 2 assumes RAS as an activator
of MEKK1. Node colours represent the activity levels $15$ minutes
after stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph2}}~
After the addition of the first hypothesis (activation of IL-1$\alpha$ production depending both
on JNK1 and ERK): production of IL-1$\alpha$ after stimulation with 100 ng/ml TNF$\alpha$ (series {\sf IL-1a~(TNFa)})
compared with stimulation with 100 ng/ml TGF$\alpha$ (series {\sf IL-1a~(TGFa)}) (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph4}}~
After the addition of the second hypothesis (activation of MEKK1 downstream of EGFR):
stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
Suppl. Sect.~\ref{suppl:parameters-tnf-egf} explains how the dosage of 5 ng/ml TNF$\alpha$ was represented in the model.\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-complete}
\end{figure*}

@@

\section*{Discussion}
\subsection*{Final remarks on the models}
We described the construction of an ANIMO
model of the circadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modelling abstractions on which ANIMO is based. The biggest
difference between the construction of these models is that the model in~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO a number of network
nodes is drawn for the molecules involved.
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter
that defines the strength of each
interaction. This approach to construct a model is closer to biological practice.
Further contributing to an interactive modelling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.
All this comes at the price of lower model precision: the curves representing
oscillation of protein activities in the ANIMO model are not as smooth
nor as precise as those obtained from the original ODE model (see Suppl. Fig.~\ref{suppl-fig:grafici-drosophila}).

We also showed the construction of an executable model
of signalling events downstream of
TNF$\alpha$ and EGF in human colon carcinoma cells. This data set has been used for
previous modelling studies, based on partial least-squares regression and fuzzy
logic~\cite{pathway-leastsquare,pathway-fuzzy}.
The partial least-squares regression model describes an abstract data-driven model
that uses statistical correlations
to relate signal transduction events to various cellular decisions. This type of modelling is
very useful in uncovering new and unexpected relations. It is also successful in making
predictions, but gives little direct insight in the dynamic behaviour of the network (see Fig.~\ref{fig:animo-spectrum}). Fuzzy
logic analysis led to a model that gave a better fit to the dynamic network behaviour than
discrete logic (Boolean) models. Inspection of the inputs to the logical gates that were used
to model protein behaviour led to the prediction of novel interactions between proteins,
showing the usefulness of this approach. For most of the proteins, such as JNK1, time was
used as an input parameter. For example, the logical gates ``if TNF$\alpha$ is high
\emph{AND} time is low, then JNK1 is high'' and ``if TNF$\alpha$ is high \emph{AND} time is
high, then JNK1 is low'' were used to
describe the dynamic behaviour of JNK1. Although this leads to a representative
description of the dynamic behaviour of JNK1, peaks in protein activity at early time points,
as measured in wet-lab experiments, were
not reproduced by the fuzzy logic model. Moreover, the fuzzy logic model gave no insight in the molecular interactions
that are involved in activation or inhibition.
